Exploring the Role of Feedback in Machine Learning Jupyter Notebooks

### 2024-06-03 report/emse24.tex
feat(report): emse24 updates

%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
% \documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage[most]{tcolorbox}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
\usepackage{wrapfig}

@@

\lstdefinestyle{mystyle}{
  language=Python,
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{delim},
  stringstyle=\color{punct},
  commentstyle=\color{numb},
  breakatwhitespace=true,
  breaklines=true,
  keepspaces=true,
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  tabsize=2,
  stepnumber=1,
  numbersep=3pt,
  captionpos=b,
  numbers=left,
  numberstyle=\tiny\color{gray},
  backgroundcolor=\color{backcolor},
  aboveskip=0pt,
  belowskip=0pt
}
\lstset{style=mystyle}

@@

\subsubsection{Motivating Example}

@@

\subsubsection{Research Questions}

@@

\todo{TODO: might need a visual aid here}

@@

\todo{TODO: point out the poor quality of notebooks out there (we have citations for this); and the resulting noise in the assertions/outputs. Thats why we need to apply these techniques to surface unique and interesting candidates!}

@@

\subsection{(RQ1) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}

\subsubsection{Data distribution check ($N = 7$)}
% (O2, O4, O9, O14, O20, O25, O48)

% TODO: we refer to these as outputs but they are actually the python statements that produce the outputs...do we need better terminology here? To me this is especially confusing because in the figure below we are actually showing the output, not the python statement (which makes sense, its more interesting to show the actual visualisation in these cases rather than the code).
\begin{table}
\centering
\caption{Examples of cell outputs used to verify distribution of data.}
\begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
\toprule
\emph{\textbf{Key}}&
\emph{\textbf{Code}}&
\emph{\textbf{Visual Output}}\\
\midrule

@@

O25&
\lstinline[]$sns.countplot(house_pred['OverallQual'])$&
\\

@@

% TODO: this is probably implications/discussions material?
Understanding the distribution of data is crucial for making informed decisions about necessary transformation steps in data analysis. For instance, visualizations can efficiently determine if scaling, normalizing, or handling outliers is needed. During the exploratory data analysis (EDA) phase, visualizations and pandas dataframes are commonly used to assess specific columns' distribution, helping identify features related to the target variable and informing feature inclusion in model training. Additionally, descriptive statistics are used post-transformation to verify the effectiveness of steps like data normalization, ensuring data conforms to the expected format for further analysis.

\subsubsection{Data relationship check ($N = 2$)}
% (O6, O10)

\begin{table}
  \centering
  \caption{Cell outputs used to verify linear relationship between features in the data.}
  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
    \toprule
    \emph{\textbf{Key}}&
    \emph{\textbf{Code}}&
    \emph{\textbf{Visual Output}}\\
    \midrule

@@

@@

% NOTE: O107 is a print statement which is execute even if the prior statement that loads the model fails...
Table~\ref{tab:resource-check} presents case studies where the output is used to verify availability of resources on the system in which the notebook is being executed. Outputs \emph{P68}, \emph{P82} and \emph{P86} check for the availability of compute resources, such as a GPU or a TPU. This study also finds outputs used to verify that a dataset or a pre-trained model has been successfully loaded into memory. For example, Output \emph{P107} is written after loading a pre-trained model is loaded into memory. Output \emph{O64} generates a pandas dataframe with the last five rows of the data while \emph{O66} shows the shape of the data. These statements are written directly after loading the data from the file system to manually verify that the data was loaded without errors. Additionally, output \emph{P71} is used to ensure that a certain version of an external library is present on the system.

The print statement P71 highlights the lack of proper software engineering best practices within Jupyter notebooks. This shortcoming in computational notebooks has been identified by prior studies~\cite{quaranta2021kgtorrent, pimentel2019large-scale}. In this case study, managing dependencies through external tools is recommended. Python's built-in \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file allows specifying dependencies and versions. Alternatively, external programs like \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/} use lockfiles to ensure specific version control, guaranteeing reproducibility across different systems.

@@

    P66&
    \lstinline[]$print('Total Run Time:')$\\
  \end{tabular}
  \label{tab:exec-time}
\end{table}

Table~\ref{tab:exec-time} presents print statement \emph{P66} which prints the total execution time of a code cell which trains an ML model with various hyper-parameter configurations using a grid-search. \emph{P66} is also used in other code cells in the notebook that train other ML models.

@@

@@

P74&
\lstinline[]$print(train_df.isnull().sum())$&
\\

@@

O36&
\lstinline[]$test.isna().sum().unique()$&
\\
\bottomrule
\end{tabular}
\label{tab:missing-value}
\end{table}

@@

@@

@@

Table~\ref{tab:network-architecture} shows print statement \emph{P92} which shows the architecture of a neural network.

@@

@@

@@

@@

% The check in P67 involves a manual verification of the output of an activation function in a neural network, which is essential for confirming the functional efficacy of neural components in model training.

@@

\subsection{(RQ2) How are assertions used to validate ML code written in Jupyter notebooks?}

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

\subsection{(RQ3) How are assertions used to validate visual outputs in Jupyter notebooks?}

\subsubsection{Model Performance}

@@

\subsubsection{Distribution Check}

@@

\subsubsection{Data Validation}

@@

\section{Implications}\label{sec:discuss}
% NOTE: from ICST paper, need to integrate...

% NOTE: breck2017ml proposes feature and data testing; we can say that the assertions in the notebook bring the system closer to migrating them into dedicated data validation systems

% NOTE: refer back to sec:assert-model-perf; the use of np.isclose and robustness to stochastic nature

% NOTE: perhaps also argue the other side, the demographic using notebooks do not have SE training; they are perhaps directly sharing the notebook and the data directly...
P71 highlights the lack of proper software engineering best practices within Jupyter notebooks. This shortcoming in computational notebooks has been identified by prior studies~\cite{quaranta2021kgtorrent, pimentel2019large-scale}. In this case study, managing dependencies through external tools is recommended. Python's built-in \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file allows specifying dependencies and versions. Alternatively, external programs like \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/} use lockfiles to ensure specific version control, guaranteeing reproducibility across different systems.

% NOTE: think about the actors: educators, researchers, developers...
The results from RQ1 provide empirical evidence to support the use of
visualisations to test various properties of ML systems.
Visualisations are an efficient and effective medium to present
complex information, that is easier to interpret by humans with the
aid of visual cues. Visualisations used to test ML system properties
however, demand manual verification effort. Information gained from
visualisations and subsequent decisions made based on this
information, must be manually verified in every iteration of the ML
development cycle.

% NOTE link to accumulation of technical debt due to highly
% experimental nature, and constant change in the system

Assertions or analytical tests extracted from ML visualisations can
reduce this manual verification effort. Assertions encapsulate the
knowledge gained from the visualisation with regards to the property
of the ML system which is currently under test. Formal assertions
document the observations made by the AI practitioner regarding the
model or the data, at a specific point in time. These observations
should not be violated by the future model development cycles unless
the requirements of the ML system are changed. Assertions scale across
organisational or personnel changes as future AI practitioners can
refer to the assertions to understand how the visualisations were
previously interpreted.

The results for RQ1 indicate that the combination of ML visualisations
with analytical assertions is an emerging testing practice within the
machine learning community. We found only $54,070$ notebooks that
contain the keyword ``assert'', while there are more than a million
Jupyter notebooks available on GitHub. This suggests that writing
assertions in computational notebooks is still relatively uncommon. We
believe that this is due to a lack of knowledge among practitioners
regarding testing approaches and the absence of adequate tooling to
support the assertion of model properties. We argue that guidelines on
ML coding practices require improvement. And advocate the need to
steer education and research of Software Engineering practices towards
the specific challenges faced when contributing to ML software
systems.

In the 1764 notebooks analysed in this study, several visualisations
lack assertions. However, implicit expectations still exist within
each visualisation which cannot be readily validated by other project
collaborators. ML testing demands proficiency in Machine Learning,
Software Testing, and the underlying Mathematical concepts. This
further highlights the need for refining the education for these
professionals and crafting tools to compensate for gaps in their
knowledge.

Results from RQ2 reveal two challenges in the current landscape of ML
testing. The assertions in the VA pairs identified in this study lack
dedicated testing libraries. This indicates that the current culture
of testing ML code is still in its infancy and ML practitioners are
not yet exploring software testing libraries. Second, the existing
testing libraries are not tuned to the specifics of ML code and
practitioners have to fall back to basic built-in approaches.

The widespread use of ``magic'' thresholds---often rooted in domain
expertise or experience obtained over time---poses risks to
maintainability of ML projects. We advocate that such thresholds are
documented for future reference. And find a clear need for automated
tools that highlight undocumented choices or suggest standardized
documentation in Markdown or code annotations.

In the results from RQ3, we see several examples of assertions that
fail to capture all the information gained from its visual
counter-part. This can lead to a mismatch of expectations in future
versions of the ML software. For instance, another project
collaborator might run the code with a new version of the dataset or a
different learning technique and assume that if the assertions run
successfully, the expected patterns from the visualisation are also
satisfied. The examples presented in Section~\ref{sec:result-rq3} show
that deriving assertions from visualisations is not trivial. Basic
visualisation patterns require complex mathematical techniques which
may be beyond the expertise of an ML developer.

For example, a linear correlation between the true and predicted
labels is evident in Figure~\ref{fig:r2}. Yet, its formal assertion in
Listing~\ref{lst:r2} necessitates defining an appropriate threshold
for the coefficient of determination metric. Our results show that
current software testing practices fail to capture the idiosyncrasies
of ML code. There is a pressing need for automated tools, that bridge
the disparity between the qualitative evaluation of ML systems using
visualisations and the extraction of analytical assertions from these
visuals.

>>>>

%
% First comes an example EPS file -- just ignore it and
% proceed on the \documentclass line
% your LaTeX will extract the file if required
\begin{filecontents*}{example.eps}
%!PS-Adobe-3.0 EPSF-3.0
%%BoundingBox: 19 19 221 221
%%CreationDate: Mon Sep 29 1997
%%Creator: programmed by hand (JK)
%%EndComments
gsave
newpath
  20 20 moveto
  20 220 lineto
  220 220 lineto
  220 20 lineto
closepath
2 setlinewidth
gsave
  .4 setgray fill
grestore
stroke
grestore
\end{filecontents*}
%
\RequirePackage{fix-cm}
%
% \documentclass{svjour3}                     % onecolumn (standard format)
%\documentclass[smallcondensed]{svjour3}     % onecolumn (ditto)
\documentclass[smallextended]{svjour3}       % onecolumn (second format)
%\documentclass[twocolumn]{svjour3}          % twocolumn
%
\smartqed  % flush right qed marks, e.g. at end of proof
%
\usepackage{graphicx}
\usepackage[most]{tcolorbox}
\usepackage{listings}
\usepackage{xcolor}
\usepackage{subcaption}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{booktabs}
\usepackage[colorinlistoftodos,prependcaption,textsize=tiny]{todonotes}
\usepackage{wrapfig}
\usepackage{framed}
\newcommand{\highlight}[1]{\begin{framed}%
  \noindent#1
\end{framed}}

@@

\lstdefinestyle{mystyle}{
  language=Python,
  basicstyle=\ttfamily\footnotesize,
  keywordstyle=\color{delim},
  stringstyle=\color{punct},
  commentstyle=\color{numb},
  breakatwhitespace=true,
  breaklines=true,
  keepspaces=true,
  showspaces=false,
  showstringspaces=false,
  showtabs=false,
  tabsize=2,
  stepnumber=1,
  numbersep=3pt,
  captionpos=b,
  % numbers=left,
  % numberstyle=\tiny\color{gray},
  % backgroundcolor=\color{backcolor},
  aboveskip=0pt,
  belowskip=0pt
}
\lstset{style=mystyle}

@@

\subsection{Motivating Example}

@@

\subsection{Exploratory Analysis}

We undertook an exploratory analysis of the collected data to formulate our research questions.

\subsection{Research Questions}

@@

% TODO: might need a visual aid here

@@

% TODO: point out the poor quality of notebooks out there (we have citations for this); and the resulting noise in the assertions/outputs. Thats why we need to apply these techniques to surface unique and interesting candidates!

@@

% TODO: some sort of introduction to the following sub-sections here; we need to explain that for code cell outputs we try to show the python statement that produces the output along with the actual output. Somethimes actual output is not possible (eg. P68) because the statement may be inside an if statement.

\subsection{(RQ1) How is feedback from the output of code cells used when writing ML code in Jupyter notebooks?}~\label{sec:result-rq1}

\subsubsection{Data distribution check ($N = 7$)}~\label{sec:data-distribution-output}
% (O2, O4, O9, O14, O20, O25, O48)

\begin{table}
  \centering
  \caption{Examples of cell outputs used to verify distribution of data.}
  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
    \toprule
    \emph{\textbf{Key}}&
    \emph{\textbf{Code}}&
    \emph{\textbf{Visual Output}}\\
    \midrule

@@

    O25&
    \lstinline[]$sns.countplot(house_pred['OverallQual'])$&
    \includegraphics[width=\linewidth]{o25.png}\\

@@

Understanding the distribution of data is crucial for making informed decisions about necessary transformation steps in data analysis. For instance, visualizations can efficiently determine if scaling, normalizing, or handling outliers is needed. During the exploratory data analysis (EDA) phase, visualizations and pandas dataframes are commonly used to assess specific columns' distribution, helping identify features related to the target variable and informing feature inclusion in model training. Additionally, descriptive statistics are used post-transformation to verify the effectiveness of steps like data normalization, ensuring data conforms to the expected format for further analysis.

\highlight{\textbf{Finding 1.1} The study identifies seven outputs used to check data distribution, including visualizations like categorical plots, KDE plots, and count plots (O2, O9 and O25), and statistical summaries from pandas dataframes (O14 and O48). These checks are used for making informed decisions about data transformations such as scaling, normalizing and handling outliers, ensuring data integrity and improving model training reliability during the EDA phase.}

\subsubsection{Data relationship check ($N = 2$)}~\label{sec:linear-relation-output}
% (O6, O10)

\begin{table}
  \centering
  \caption{Cell outputs used to verify linear relationship between features in the data.}
  \begin{tabular}{@{}m{0.05\textwidth} m{0.4\textwidth} m{0.4\textwidth}@{}}
    \toprule
    \emph{\textbf{Key}}&
    \emph{\textbf{Code}}&
    \emph{\textbf{Output}}\\
    \midrule

@@

\highlight{\textbf{Finding 1.2} The study identifies two outputs used to verify linear relationships between features in the data. Visualizations such as a scatter plot with a regression line (O10) and a line plot (O6) are employed to assess the linearity between features. These checks are used for identifying and potentially removing redundant features or performing feature selection to ensure that the training dataset is optimized for linear ML models.}

@@

% NOTE: O107 is a print statement which is execute even if the prior statement that loads the model fails...
Table~\ref{tab:resource-check} presents outputs used to verify availability of resources on the system in which the notebook is being executed. Outputs \emph{P68}, \emph{P82} and \emph{P86} check for the availability of compute resources, such as a GPU or a TPU. This study also finds outputs used to verify that a dataset or a pre-trained model has been successfully loaded into memory. For example, Output \emph{P107} is written after loading a pre-trained model is loaded into memory. Output \emph{O64} generates a pandas dataframe with the last five rows of the data while \emph{O66} shows the shape of the data. These statements are written directly after loading the data from the file system to manually verify that the data was loaded without errors.

Additionally, print statement \emph{P71} is used to ensure that a certain version of an external library is present on the system. This is interesting since external dependencies in Python projects are typically managed using dependency management solutions such as a \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file or external tools such as \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/}.

\highlight{\textbf{Finding 1.3} The study identifies seven outputs used to verify resource availability on the systems where the Jupyter notebooks are being executed. These checks include confirming the availability of compute resources like GPUs and TPUs (P68, P82 and P86), ensuring datasets or pre-trained models are successfully loaded (P107, O64 and O66), and verifying the presence of specific library versions (P71). These checks are used for ensuring that the computational environment is correctly set up, preventing execution errors, and facilitating smooth workflow execution.}

@@

    P66&
    \lstinline[]$print('Total Run Time:')$\\
    \bottomrule
  \end{tabular}
  \label{tab:exec-time}
\end{table}

Table~\ref{tab:exec-time} presents print statement \emph{P66} that prints the total execution time of a code cell which trains an ML model with various hyper-parameter configurations using a grid-search. \emph{P66} is also used in other code cells in the notebook that train other ML models.

@@

\highlight{\textbf{Finding 1.4} The study identifies one output (P66) used to check the total execution time of training and cross-validating an ML model. The print statement is used for evaluating the efficiency of model training processes, especially when multiple models are trained iteratively.}

@@

    P74&
    \lstinline[]$print(train_df.isnull().sum())$\\

@@

    O36&
    \lstinline[]$test.isna().sum().unique()$\\
    \bottomrule
  \end{tabular}
  \label{tab:missing-value}
\end{table}

@@

\highlight{\textbf{Finding 1.5} The study identifies three outputs (P74, O12 and O36) used to check for missing data. Missing values can lead to biased or incorrect conclusions, and many ML algorithms require complete numerical datasets for accurate calculations. Identifying and addressing missing data during preprocessing is essential to avoid errors and improve model reliability.}

@@

\highlight{\textbf{Finding 1.6} The study identifies 33 outputs used to check the performance of trained ML models. These include outputs that print key performance metrics (P3, P6, P18, P50), a normalised heatmap used to visualise the confusion matrix of a model's predictions (O3) and a custom function written for manual evaluation of model predictions on random input samples (O52). These checks are used throughout the iterative model development phase, enabling practitioners to compare different models or variations, adjust parameters, and refine features to improve performance.}

@@

% TODO: need more description and context around this print statement (where was it written, and why?)
Table~\ref{tab:network-architecture} shows print statement \emph{P92} which shows the architecture of a neural network.

@@

\highlight{\textbf{Finding 1.7} The study identifies one output (P92) used to print the architecture of a neural network. Such outputs can be used during the model development phase to ensure compatibility between the layers of a neural network, for understanding and learning about the architecture of pre-trained models and for documentation purposes.}

@@

\highlight{\textbf{Finding 1.8} The study identifies three outputs used to check the shape of data. These checks are used to ensure data dimensions align with expectations, particularly after preprocessing or transformation steps.}

@@

\highlight{\textbf{Finding 1.9} The study identifies two outputs used to verify the data types of features in the dataset. These checks are used to determine the necessary preprocessing or transformation steps requried to make the data suitable for training. Type checks are also used to ensure that all features are numerical prior to performing mathematical operations such as matrix multiplication, which is used in several ML models.}

@@

\highlight{\textbf{Finding 1.10} This study finds five outputs used to perform spot checks at various stages of the MLDL. Spot checks are used to ensure data integrity by verifying data transformation and feature engineering steps meet expectations.}

@@

\highlight{\textbf{Finding 1.11} The study identifies four outputs used to monitor model training progress. These outputs provide real-time feedback on training loss and accuracy (O8), validate model adherence to expected data patterns (O31), and identify optimal parameters through tuning methods (O42). These checks facilitate timely adjustments during the training process to prevent overfitting and ensure model performance optimization.}

\subsection{(RQ2) How are assertions used to validate ML code written in Jupyter notebooks?}~\label{sec:result-rq2}

@@

\highlight{\textbf{Finding 2.1} This study identifies three assertions validating batch size configurations. These assertions ensure batch sizes are divisible by dataset sizes, image dimensions are appropriately divided by patch sizes, and the number of images exceeds the batch size. These checks are crucial for optimizing computational efficiency and ensuring stable and smooth model training.}

@@

\highlight{\textbf{Finding 2.2} This study identifies one assertion to ensure there are no overlapping values between training and validation sets to prevent data leakage. Preventing data leakage is essential to avoid overfitting and ensure the model's ability to generalize to new data. This practice maintains the integrity and validity of the model evaluation metrics.}

@@

\highlight{\textbf{Finding 2.3} Eight assertions ensure the presence of specific columns in the dataset and the absence of missing values. These checks prevent operations on empty datasets and confirm the absence of \texttt{NaN} values, maintaining data integrity and reliability for model training.}

@@

\highlight{\textbf{Finding 2.4} Four assertions validate key mathematical properties of data and model outputs. These include ensuring dimensional consistency in convolution operations, monitoring standard deviation for batch normalization suitability, verifying uniformity in computations, and checking matrix symmetry. These checks confirm the robustness and reliability of model computations.}

@@

\highlight{\textbf{Finding 2.5} Eleven assertions evaluate model performance against predefined thresholds for accuracy, precision, recall, F1 score, and other metrics. These checks validate the model's predictive accuracy and reliability, ensuring it meets expected performance standards despite the stochastic nature of ML models.}

@@

\highlight{\textbf{Finding 2.6} Three assertions verify the consistency of neural network architecture. These include ensuring weight dimensions match across layers and that activation shapes and regularization parameters are as expected. These checks ensure that the neural network is correctly configured and integrated with pre-trained models, maintaining the integrity of data flow during training.}

@@

\highlight{\textbf{Finding 2.7} Seven assertions confirm the availability of essential resources like pre-trained models, data files, successful model loading, and completion of training processes. These checks prevent runtime errors and ensure proper code execution, maintaining the reliability and robustness of the ML workflows.}

@@

\highlight{\textbf{Finding 2.8} 26 assertions verify the dimensions of various data structures such as input features, labels, predictions, images, sequences, and text embeddings. These checks ensure data consistency and compatibility throughout the ML pipeline, preventing runtime errors and ensuring accurate model training and evaluation.}

@@

\highlight{\textbf{Finding 2.9} Five assertions validate the data types of input features, models, and intermediate objects. These checks ensure compatibility with ML models and operations, preventing errors from incompatible methods or attributes, thus maintaining the reliability of the ML workflow.}

@@

\highlight{\textbf{Finding 2.10} Fourteen assertions perform validation checks on data structures, ensuring the presence of specific values, verifying value ranges, and confirming uniqueness or cardinality of values. These checks catch potential errors or inconsistencies early, ensuring data meets specific criteria or constraints for robust ML applications.}

\subsection{(RQ3) How are assertions used to validate visual outputs in Jupyter notebooks?}~\label{sec:result-rq3}

\subsubsection{Model Performance ($N = 15$)}

@@

\highlight{\textbf{Finding 3.1} 15 VA pairs are used to evaluate the performance of ML models. The visualizations include metrics such as mean reward, training loss, accuracy, BLEU scores, and R2 scores. Assertions associated with these visualizations ensure models meet a specific performance criteria.}

\subsubsection{Distribution Check ($N = 1$)}

@@

\highlight{\textbf{Finding 3.1} One visualization is used to display the distribution of bootstrapped sample means. The corresponding assertion checks if the bootstrapped mean lies between a specific range, ensuring a consistent statistical estimation.}

\subsubsection{Data Validation ($N = 9$)}

@@

\highlight{\textbf{Finding 3.3} Nine VA pairs focus on data validation tasks such as comparing the original with the scaled data distribution and ensuring augmented data is different from the original. These validations confirm that data preprocessing steps like scaling, transformation, and augmentation are correctly applied.}

\section{Discussion}\label{sec:discuss}

Our results from RQ1 reveal that outputs from code cells in Jupyter notebooks are extensively used to make design and implementation decisions during the ML development cycle. Making such decisions however, requires manual validation of the outputs. For instance, we find several instances where practitioners use visualisations to check data distributions and print statements to verify data shapes and types before proceeding with model training. Critical decisions, such as deciding whether to normalize features based on their distributions or adjusting model parameters based on training feedback, rely heavily on these manual checks. Like traditional software systems, an ML pipeline is subject to change due to evolving model requirements, personnel turnover, and changes in the data itself. Such changes can introduce inconsistencies or errors that manual checks might miss, highlighting the need for more automated validation mechanisms.

Assertions can play a crucial role in automatically validating critical assumptions in ML workflows. As seen in the results from RQ2, assertions can verify that data distributions remain consistent with expectations or that model performance metrics meet predefined thresholds. Embedding assertions into the code helps ensure that any deviation from expected conditions is promptly flagged, reducing the reliance on manual verification and enhancing the robustness and maintainability of ML pipelines. Assertions also serve as an executable form of documentation that captures the assumptions and decisions made during model development. This is particularly important in collaborative environments where multiple practitioners might work on the same project. Assertions ensure that all team members have a clear understanding of the requirements and constraints across the entire ML pipeline, facilitating smoother transitions and handovers. Moreover, they enhance the reproducibility of ML experiments by ensuring that the same validation checks are applied consistently, regardless of the execution order of the notebook cells.

% P71 highlights the lack of proper software engineering best practices within Jupyter notebooks. This shortcoming in computational notebooks has been identified by prior studies~\cite{quaranta2021kgtorrent, pimentel2019large-scale}. In this case study, managing dependencies through external tools is recommended. Python's built-in \emph{requirements.txt}~\footnote{https://pip.pypa.io/en/stable/reference/requirements-file-format/} file allows specifying dependencies and versions. Alternatively, external programs like \emph{pipenv}~\footnote{https://pipenv.pypa.io/en/latest/} and \emph{poetry}~\footnote{https://python-poetry.org/} use lockfiles to ensure specific version control, guaranteeing reproducibility across different systems.

% NOTE: here we need to bring in prior literature on ML testing; and how majority of the research is working on testing using a test dataset; however we also have data and models
% NOTE: 3 pillars of of ML (sato2019continuous)

From the 3M Jupyter notebooks collected in this study, only 30K (9.7\%) were found to utilize assertions. This indicates that while the use of assertions in Jupyter notebooks is still uncommon, it represents an emerging testing practice within the ML community. The relatively low adoption rate suggests that there is significant room for growth and improvement in the integration of assertions into ML development workflows.

We believe this limited use of assertions is due to a lack of knowledge among practitioners regarding testing approaches and the absence of adequate tooling to support the assertion of data and model properties. Current guidelines on ML coding practices require enhancement to emphasize the importance of testing. There is a pressing need to steer education and research in Software Engineering towards addressing the specific challenges faced when developing ML-enabled software systems. Fostering a deeper understanding of testing practices and providing tools tailored to ML development can help promote more rigorous and reliable testing methodologies within the community.

The assertions presented in RQ2 and RQ3 further reveal two significant challenges in the current landscape of testing in ML Jupyter notebooks. First, the assertions identified in this study lack the support of dedicated testing libraries. This suggests that the culture of testing ML code is still in its infancy, and ML practitioners are not yet fully exploring or leveraging software testing libraries. Secondly, existing testing libraries are not specifically tuned to the needs of ML code, leading practitioners to rely on basic built-in approaches. This highlights the necessity for developing specialized testing frameworks that cater to the unique requirements of ML projects.

Visualizations are an effective tool for interpreting complex ML models and data transformations. However, the manual effort required to validate insights derived from these visualizations can be substantially more compared to text output such as the ones presented in Section~\ref{sec:result-rq2}. We argue that the need for translating visual insights into formal assertions is more pressing. The visual cue and trends observed in visualisations are open to iterpretation, thus the interpretation of the same plot may differ amongst two practitioners. Translating visual insights into assertions ensures taht the knowledge encapsulated in the visualisations along with its interpretation and subsequent decisions made based on this interpretation is preserved and consistently applied throughout the ML lifecycle.

In the VA pairs from RQ3, we see several examples of assertions that fail to capture all the information gained from its visual counter-part. This can lead to a mismatch of expectations in future versions of the ML software. For instance, another project collaborator might run the code with a new version of the dataset or a different learning technique and assume that if the assertions run successfully, the expected patterns from the visualisation are also satisfied. The VA pairs presented in Section~\ref{sec:result-rq3} show that deriving assertions from visualisations is not trivial. Basic visualisation patterns require complex mathematical techniques which may be beyond the expertise of an ML developer. For example, a linear correlation between the true and predicted labels is evident in VA107 presented in Table~\ref{tab:va-model-perf}. Yet, its formal assertion necessitates defining an appropriate threshold for the coefficient of determination metric. Our results show that current software testing practices fail to capture the idiosyncrasies of ML code. There is a pressing need for automated tools, that bridge the disparity between the qualitative evaluation of ML systems using visualisations and the extraction of analytical assertions from these visuals.

% TODO: integrate our NLBSE paper
Our results show a clear need for developing ML-specific testing tools. Such tools should seamlessly integrate with computational notebooks and provide automated suggestions for assertions can greatly enhance the reliability and robustness of ML models. These tools can also aid in documenting and standardizing the use of domain-specific "magic" thresholds, ensuring that they are consistently applied and maintained.

% NOTE: we find overlap between the manual tests performed using the output of code cells and the assertions.

% NOTE: bring in prior work on reproducibility, code quality here
% NOTE: notebooks have been adopted by a much larger scientific community; not eveybody has a background in software engineering; we need more tools to aid notebook users...

\section{Implications}

\paragraph{\textbf{For ML developers}} ML developers are strongly encouraged to incorporate assertions into their Jupyter notebooks. Assertions play a crucial role in validating critical assumptions automatically, reducing the reliance on manual checks, and enhancing the robustness and maintainability of ML pipelines. By embedding assertions, developers can document the decisions made during model development, ensuring that these decisions are consistently applied and preserved throughout the ML lifecycle. This practice not only facilitates smoother transitions and handovers in collaborative environments but also improves the reproducibility of ML experiments by ensuring consistent validation checks.

\paragraph{\textbf{For Researchers}} The use of assertions in Jupyter notebooks, though currently uncommon, is an emerging testing practice within the ML community. This study highlights the potential for assertions to enhance the reliability and robustness of ML models by automating the validation of critical assumptions. Researchers are encouraged to explore the development and evaluation of ML-specific testing methodologies that leverage assertions. This could involve creating new testing frameworks and tools tailored to the unique challenges of ML systems, thus contributing to the advancement of software engineering practices in the ML domain.

\paragraph{\textbf{For Notebook Tool Developers}} There is a clear need for automated tools that suggest assertions based on the outputs of code cells in Jupyter notebooks. Tool developers can play a pivotal role by creating ML-specific testing tools that seamlessly integrate with computational notebooks. These tools should provide automated suggestions for assertions, helping practitioners to translate visual insights into formal, executable checks. By bridging the gap between qualitative evaluation and analytical assertions, such tools can significantly enhance the reliability and robustness of ML models, ensuring that they meet predefined performance standards consistently.

\paragraph{\textbf{For Educators}} Educators should emphasize the importance of testing ML code written in Jupyter notebooks as part of the curriculum. By incorporating testing practices and the use of assertions into ML education, students can develop a deeper understanding of the importance of automated validation in ensuring model reliability and robustness. Teaching these best practices early on can help cultivate a culture of rigorous testing among future ML practitioners, preparing them to address the specific challenges faced in ML-enabled software systems.

\paragraph{\textbf{For General Notebook Users}} Awareness of software engineering best practices, including the use of assertions, is crucial for all users of Jupyter notebooks. Assertions serve as an executable form of documentation, capturing the critical assumptions and decisions made during the development process. By adopting assertions, users can ensure that their notebooks are more robust, maintainable, and reproducible. This practice helps in maintaining consistency and reliability, especially in collaborative projects where multiple users might work on the same notebook, ensuring that the assumptions and validations are transparent and consistently applied.

\section{Future Work}

\section{Conclusion}
